Finite-state morphological transducers for three Kypchak languages

### 2013-10-24 kipchak-lrec-2013/kipchak-paper.tex
some citation fixes, etc.
git-svn-id: https://svn.code.sf.net/p/apertium/svn/branches/papers/@48179 72bbbca6-d526-0410-a7d9-f06f51895060

\cite{bekmanova2013}

@@

\begin{map*}[htbp]
	\includegraphics[width=\textwidth]{map/map}
	\caption{The three branches of Kypchak (N, S, W), showing the distribution of the three languages for which transducers were developed (tat, kaz, kum).  Language codes are from ISO 639-3.}
	\label{map1}
\end{map*}

@@

\subsection{Kazakh}
Kazakh /q{\symbl ɑ}z{\symbl ɑ}q/ is spoken primarily in Kazakhstan, where it is the national language, sharing official status with Russian as an official language.  Large communities of native speakers also exist in China, neighbouring Central-Eurasian republics, and Mongolia.  Estimates of the total number of speakers range from 8 million \citep{ethnologue} to 11 million \citep{encyclopedin} people.

@@

\cite{bammatov1960}
\cite{olmesov2000}

\section{Background}
\subsection{Morphological transducers}
%FIXME: need a source, maybe something by Tommi?
The objective of a morphological transducer is twofold: firstly to take surface forms (e.g., алдым) and generate all possible lexical forms, and secondly to take lexical forms (e.g.,  ал{\tt {\small <v><tv><ifi><p1><sg>}}, алд{\tt {\small <n><px1sg><nom>}}, etc.) and generate one or more surface forms.  Since they are implemented as finite-state automata \citet{FSTs}, they are reversible by default.

The transducers were designed based on the Helsinki Finite State Toolkit \citep{linden2011} which is a free/open-source reimplementation of the Xerox finite-state toolchain, popular in the field of morphological analysis.  It implements both the \textbf{lexc} formalism for defining lexicons, and the \textbf{twol} and \textbf{xfst} formalisms for modeling morphophonological rules.  It also supports other finite state transducer formalisms such as \textbf{sfst}.  This toolkit has been chosen as it -- or the equivalent XFST -- has been widely used for other Turkic languages, such as Turkish \citep{coltekin2010}, Crimean~Tatar~\citep{altintas2001}, Turkmen \citep{tantug2006}, and Kyrgyz \citep{washington2012}, and is available under a free/open-source licence.

@@

As the authors found it difficult to locate native speakers of Kumyk, the morphotactics of the Kumyk transducer were developed based on the existing encoded morphotactics of Kazakh and Tatar (and occasionally Kyrgyz), with consultation of corpora, as described above.

@@

\begin{table}
\begin{center}
\begin{tabular}{lrrr}
		\toprule
\multirow{2}{*}{\textbf{Part of speech}} & \multicolumn{3}{c}{\textbf{Number of stems}} \\ \cline{2-4}
                        & Kazakh & Tatar & Kumyk \\
		\midrule
		Noun & - & - & - \\
		Verb & - & - & - \\
		Adjective & - & - & - \\
		Proper noun & - & - & - \\
		Adverb & - & - & - \\
		Numeral & - & - & - \\
		Conjunction & - & - & - \\
		Postposition & - & - & - \\
		Pronoun & - & - & - \\
		Determiner & - & - & - \\
		\midrule
		Total: & - & - & - \\
		\bottomrule
\end{tabular}
 \caption{Number of stems in each of the categories}
 \label{table:coverage}
\end{center}

>>>>

@@

\begin{map*}[htbp]
	\includegraphics[width=\textwidth]{map/map}
	\caption{The three subbranches of Kypchak (N, S, W), roughly divided with black lines, showing the geographic distribution of the three languages for which transducers were developed (tat, kaz, kum).  Language codes are from ISO 639-3.}
	\label{map1}
\end{map*}

@@

\subsection{Kazakh}
Kazakh /q{\symbl ɑ}z{\symbl ɑ}q/ is spoken primarily in Kazakhstan, where it is the national language, sharing official status with Russian as an official language.  Large communities of native speakers also exist in China, neighbouring Central-Eurasian republics, and Mongolia.  Estimates of the total number of speakers range from 8 million \citep{ethnologue} to 11 million \citep{nationalencyklopedin} people.

@@

\section{Background}
\subsection{Morphological transducers, previous work}
%FIXME: need a source, maybe something by Tommi?
The objective of a morphological transducer is twofold: firstly to take surface forms (e.g., алдым) and generate all possible lexical forms, and secondly to take lexical forms (e.g.,  ал{\tt {\small <v><tv><ifi><p1><sg>}}, алд{\tt {\small <n><px1sg><nom>}}, etc.) and generate one or more surface forms.  Since they are implemented as finite-state automata, they are reversible by default.

The transducers were designed based on the Helsinki Finite State Toolkit \citep{hfst/2011} which is a free/open-source reimplementation of the Xerox finite-state toolchain, popular in the field of morphological analysis.  It implements both the \textbf{lexc} formalism for defining lexicons, and the \textbf{twol} and \textbf{xfst} formalisms for modeling morphophonological rules.  It also supports other finite state transducer formalisms such as \textbf{sfst}.  This toolkit has been chosen as it -- or the equivalent XFST -- has been widely used for other Turkic languages, such as Turkish \citep{coltekin2010}, Crimean~Tatar~\citep{altintas2001}, Turkmen \citep{tantug2006}, and Kyrgyz \citep{washington2012}, and is available under a free/open-source licence.

The authors learned of another Kazakh morphological transducer in existence \citep{bekmanova2013} only after production-ready version of our transducer was released.  We have not yet been able to evaluate this system or compare it to ours.

@@

As the authors found it difficult to locate native speakers of Kumyk, the morphotactics of the Kumyk transducer were developed based on the existing encoded morphotactics of Kazakh and Tatar (and occasionally Kyrgyz), with consultation of corpora, as described above.  A dictionary \citep{bammatov1960} and a grammar \citep{olmesov2000} of Kumyk were also consulted as needed.

@@

\begin{table}
\begin{center}
\begin{tabular}{lrrr}
		\toprule
\multirow{2}{*}{\textbf{Part of speech}} & \multicolumn{3}{c}{\textbf{Number of stems}} \\ \cline{2-4}
                        & Kazakh & Tatar & Kumyk \\
		\midrule
		Noun & - & - & - \\
		Verb & - & - & - \\
		Adjective & - & - & - \\
		Proper noun & - & - & - \\
		Adverb & - & - & - \\
		Numeral & - & - & - \\
		Conjunction & - & - & - \\
		Postposition & - & - & - \\
		Pronoun & - & - & - \\
		Determiner & - & - & - \\
		\midrule
		Total: & - & - & - \\
		\bottomrule
\end{tabular}
 \caption{Number of stems in each of the categories}
 \label{table:stems}
\end{center}
